Cover letter for the original IJRR submission

### 2017-04-20 body/appendix_hpp.tex

\section{Source code of our planner}
\label{app:hpp}
Our planner is implemented using the Humanoid Path Planner (HPP) software, introduced in~\cite{7759083}.
HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS, that can be found at \url{http://projects.laas.fr/gepetto/index.php/Software/Main}.
The HPP modules implement the standard tools and algorithms used in motion planning,
such as the Bi-RRT planner from which RB-RRT is derived.

The robot models used in our experiments are described using the standard urdf file format, that is 
compatible with HPP.

>>>>

\section{Source code of our planner}
\label{app:hpp}
Our planner is implemented using the Humanoid Path Planner (HPP) software, introduced in~\cite{7759083}.
HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS, that can be found at \url{http://projects.laas.fr/gepetto/index.php/Software/Main}.
HPP implements the standard tools and algorithms used in motion planning,
such as the Bi-RRT planner from which RB-RRT is derived.

The robot models used in our experiments are described using the standard urdf file format, compatible with HPP.

### 2017-04-20 body/conclusion.tex

These results clearly justify the pragmatic decisions of this paper.

@@

While we have exhibited complete multi-contact locomotion obtained with our contact planner, our main concern for future work is to address the interpolation
between contact sequences ($\mathcal{P}_3$), which remains an open issue in highly-constrained scenarios.
Solving $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem in the interpolation phase, an issue 
not addressed by existing frameworks. We aim at providing our plans with transition certificates, that would define constraints on $\mathcal{P}_3$, under which
the transition between two contact configurations is feasible and collision-free.
%~ Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
%~ at the center of mass of the robot.
 %~ We are also
%~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
Finally, we aim at performing kinodynamic planning to remove the constraint that configurations in the contact plan be in static equilibrium. We believe that the most promising direction in this regard is to integrate the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
Addressing these two issues is essential to bridge the gap between the planning and control aspects of legged locomotion.

>>>>

These results clearly justify our pragmatic approach.

@@

While we have exhibited complete multi-contact locomotion obtained with our contact planner, our main concern for future work is to address the interpolation
between contact sequences ($\mathcal{P}_3$), which remains an open issue in highly-constrained scenarios.
Solving $\mathcal{P}_3$ requires addressing efficiently the collision avoidance problem in the interpolation phase, an issue 
not addressed by existing frameworks. We aim at providing our plans with transition certificates, that would define constraints on $\mathcal{P}_3$, under which
the transition between two contact configurations is feasible and collision-free.
%~ Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
%~ at the center of mass of the robot.
 %~ We are also
%~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
Finally, we aim at performing kinodynamic planning to remove the constraint that configurations be in static equilibrium. We believe that the most promising direction in this regard is to integrate the notion of Admissible Velocity Propagation \citep{DBLP:conf/rss/PhamCN13}.
Addressing these two issues is essential to bridge the gap between the planning and control aspects of legged locomotion.

### 2017-04-20 body/efort.tex
reviewer 2

\subsubsection{The force and velocity ellipsoids:}

@@

We consider: a limb configuration $\mathbf{q}^k$; its end effector position $\mathbf{p}^k$; its Jacobian matrix
$\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
Yoshikawa~\cite{Yoshikawa1984} defines the velocity~\ref{eq:vel} and force~\ref{eq:vel} ellipsoids:

@@

\subsubsection{Manipulability-based heuristics:}
From these definitions, we can derive three useful heuristics, that all account for the environment and the task being performed.
The first one, EFORT, was introduced by \cite{Tonneau2014}; the other two are new minor contributions, derived from these previous works.

>>>>

\subsubsection{The force and velocity ellipsoids}

@@

We consider: a limb configuration $\mathbf{q}^k$; its end effector position $\mathbf{p}^k$; its Jacobian matrix
$\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
Yoshikawa~\cite{Yoshikawa1984} defines the velocity(~\ref{eq:vel}) and force~(\ref{eq:for}) ellipsoids:

@@

\subsubsection{Manipulability-based heuristics}
From these definitions, we can derive three useful heuristics, that all account for the environment and the task being performed.
The first one, EFORT, was introduced by \cite{Tonneau2014}; the other two are new minor contributions, derived from these previous works.

### 2017-04-20 body/equilibrium.tex

\subsubsection{Conditions for static equilibrium:}
We first define the variables of the problem, for $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
%~ \item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
\item $m \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact point $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{\boldsymbol\upgamma}_{i1},\mathbf{\boldsymbol\upgamma}_{i2}$ form a local Cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{\boldsymbol\upgamma}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}

@@

\subsubsection{Formulation of a robust LP:}
Let $b_0 \in \mathbb{R}$ be a scalar value. We now define the following LP:

>>>>

\subsubsection{Conditions for static equilibrium}
We first define the variables of the problem, for $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
%~ \item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
\item $m \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact point $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{\boldsymbol\upgamma}_{i1},\mathbf{\boldsymbol\upgamma}_{i2}$ form a local Cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{\boldsymbol\upgamma}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}

@@

\subsubsection{Formulation of a robust LP}
Let $b_0 \in \mathbb{R}$ be a scalar value. We now define the following LP:

### 2017-04-20 body/intro.tex

Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorics. Two main directions have been explored. \textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}. A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}. To get reasonable computation times, the method uses a simplified dynamic model for the avatar. Still, the method is far from real-time  (about 1 minute of computation for 20 contacts).  A similar approach has been considered for manipulation by \cite{gabicciniisrr15}. Deits and Tedrake proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered, and equilibrium is not considered~\cite{DBLP:conf/humanoids/DeitsT14}. 
Dai et al.\cite{dai2014whole} extended the work of Posa et al.\cite{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
the contacts manually for complex interactions.
In addition to the limits of the current implementations, optimization-based approaches only converge locally.

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling has been shown by Escande et al.\cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e. an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, 
but at the cost of prohibitive computation times (up to several hours) for constraining scenarios. This approach is suboptimal because the quality of the motion depends on the quality of the guide path. Bouyarmane et al.~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium. Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented). Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path. Chung and Khatib~\cite{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment. This planning phase does not account for collisions, implying that replanning is required in case of failure. In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}.

>>>>

Most of the papers that followed the work of Bretl have explored alternative formulations to handle the combinatorics. Two main directions have been explored. \textbf{On the one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}. A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}. To get reasonable computation times, the method uses a simplified dynamic model for the avatar. Still, the method is far from real-time  (about 1 minute of computation for 20 contacts).  A similar approach has been considered for manipulation by \cite{gabicciniisrr15}. Deits and Tedrake proposed to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered, and equilibrium is not considered~\cite{DBLP:conf/humanoids/DeitsT14}. 
Dai et al.~\cite{dai2014whole} extended the work of Posa et al.~\cite{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
the contacts manually for complex interactions.
In addition to the limits of the current implementations, optimization-based approaches only converge locally.

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling has been shown by Escande et al. \cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e. an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, 
but at the cost of prohibitive computation times (up to several hours) for constraining scenarios. This approach is suboptimal because the quality of the motion depends on the quality of the guide path. Bouyarmane et al. ~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium. Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented). Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path. Chung and Khatib~\cite{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment. This planning phase does not account for collisions, implying that replanning is required in case of failure. In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}.

### 2017-04-20 body/rbprm.tex

This Section has presented a guide path planner for the geometric root of a robot. This planner is implemented as a low-dimensional sampling-based 
algorithm. Given start and goal configurations, it outputs a continuous path for the root of the robot. 
%~ \deladp{Thanks to the \textit{reachability condition}, we assume that for any configuration in the path there exists a joint configuration that results in static equilibrium.}
%~ Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
%~ By doing this we can solve the problem with a sampling-based approach, in \gls{interactive} computation times.

>>>>

This Section has presented a guide path planner for the geometric root of a robot, implemented as a low-dimensional sampling-based 
algorithm. Given start and goal configurations, it outputs a continuous path for the robot's root. 
%~ \deladp{Thanks to the \textit{reachability condition}, we assume that for any configuration in the path there exists a joint configuration that results in static equilibrium.}
%~ Thanks to these modifications, the problem of planning a \gls{contact reachable} path is reduced to a geometric collision-checking problem, of low dimension (6 for HyQ, 8 for HRP-2 that has 2 joints in the torso).
%~ By doing this we can solve the problem with a sampling-based approach, in \gls{interactive} computation times.
